---
execute:
  echo: true
format:
  html:
    suppress-bibliography: true
---

# Overview {.unnumbered}

::: {.callout-note icon="false"}
## This repository contains all data, analysis code and documentation for the manuscript:

D-alanine Aminotransferase (Dat) promotes Staphylococcus aureus colonization fitness on human nasal respiratory epithelium
:::

## Reproducibility

All analyses are provided as individual Quarto (`.qmd`) notebooks, which can be rendered independently. Each notebook reads processed `.rds` files from the `dataframes/` folder and generates publication-quality figures and statistical outputs (Table S2).

To reproduce the analyses:

-   Clone or download this [repository](https://github.com/KLemonLab/DatCol_Manuscript) to your local machine.
-   Download the `dataframes/` and `Tn-Seq/` folders and place it in your working directory.
-   Download all `.qmd` notebooks from the repository and keep them in the same directory.
-   Install required R packages listed in [R Session Info](Manuscript_DatCol_RSession.html) using `renv` or manually.
-   Render the notebooks using Quarto to generate figures and results.

You can set environment variable as follows in R:

```{r, eval = FALSE}
Sys.setenv(BOX_PATH = "/path/to/your_project")
```

Alternatively, you can add this to your `.Renviron` file:

```{r, eval = FALSE}
BOX_PATH=/path/to/your_project
```

The whole project can be rendered using RStudio's "Render" or "Build" options or in the terminal via:

```{bash, eval = FALSE}
quarto render
```

## Repository Structure

This repository is organized as follows:

```         
DatCol_project/
├── dataframes/          
│   ├── [input data / CSV tables]
│   ├── [input data / RDS files]
│   └── …
├── Tn-Seq/
│   ├── RefSeqs/
│   │   ├── [reference sequence files]
│   │   └── …
│   ├── sites/
│   │   ├── [insertion site files]
│   │   └── …
│   ├── scripts/
│   │   └── TnGeneBin.pl
└── [analysis notebooks saved as .qmd files]
```

### Data folders

-   [dataframes](https://github.com/KLemonLab/DatCol_Manuscript/tree/main/dataframes)/
    -   Contains all processed data used to generate figures and statistical analyses (see Table S2)
    -   Files are provided in `.csv` and `.rds` formats for easy visualization and reproducibility in R
    -   `.rds` file names correspond to their associated `.qmd` analysis notebooks
    -   `DatCol_CFUs_HNO_Chemical.rds` and `DatCol_CFUs_HNO_Genetic.rds` both correspond to the *Complementation* notebook
    -   `DatCol_CFUs_HNO_Controls.rds` contains all monocolonization control data associated with `HNO_Competition.qmd`
-   [Tn-Seq](https://github.com/KLemonLab/DatCol_Manuscript/tree/main/Tn-Seq)/
    -   Raw FASTQ files are available through the SRA \[insert link or BioProject accession\]
    -   Processed insertion site files are included here to reproduce the Essential Gene Analysis and associated figure generation
    -   Subfolders include:
        -   `RefSeqs/`: reference genome files
        -   `sites/`: transposon insertion site files
        -   `scripts/`: includes the `TnGeneBin.pl` Perl script used to bin insertion sites for essential gene analysis

### Analysis Notebooks

-   Main Analysis Notebooks:
    -   [Tn-Seq Pipeline](Manuscript_DatCol_Tn-Seq_Pipeline.html)
    -   [Tn-Seq Essential Genes Analysis](Manuscript_DatCol_Tn-Seq_Figures.html)
    -   [Competition Assays on HNO-ALI](Manuscript_DatCol_HNO_Competition.html)
    -   [Complementation Assays on HNO-ALI](Manuscript_DatCol_HNO_Complementation.html)
    -   [Growth Curves (OD600)](Manuscript_DatCol_Growth_Curves.html)
    -   [Cocolonization Assays with *Corynebacterium pseudodiphtheriticum*](Manuscript_DatCol_HNO_Cps.html)
    -   [Cytotoxicity Assays (LDH)](Manuscript_DatCol_LDH.html)
    -   [Alanine in HNO-ALI mucus](Manuscript_DatCol_HNO_Alanine.html)
-   Other Notebooks:
    -   [Linear Mixed Model Workflow](Manuscript_DatCol_LMMStats.html)
    -   [R Session Info](Manuscript_DatCol_RSession.html)
    -   [References](references.html)
