---
format:
  html:
    suppress-bibliography: true
---

# Linear Mixed Model Workflow {.unnumbered}

```{r}
# functions
source("R/LMM_functions.R")
```

This notebook documents the shared Linear Mixed-effects Model (LMM) workflow used throughout this repository. Each analysis notebook focuses on dataset-specific preprocessing, visualization, and interpretation, while calling shared functions to perform the statistical analysis.

The shared functions used throughout the repository and defined in `R/LMM_functions.R` are:

-   `mixed_model_with_emmeans()`
-   `format_stats_results()`

## `mixed_model_with_emmeans()`

This function fits a linear mixed-effects model and returns a structured list containing model summaries, ANOVA results, estimated marginal means, and post hoc contrasts.

### Inputs

-   `input_data`: data frame containing the variables used in the model
-   `response`: name of the response variable as a character string
-   `log_response`: logical; if `TRUE`, the response is modeled as `log(response)`
-   `fixed_effects`: character vector of fixed-effect variable names; if more than one fixed effect is provided, interaction terms are included in the model
-   `random_effects1`: first random effect, typically a grouping factor such as `line`
-   `random_effects2`: optional second random effect, typically `date`, used to define an additional random grouping term (`random_effects1:random_effects2`) and as a fallback random effect if singularity is detected
-   `adjust`: p-value adjustment method passed to `emmeans` for multiple comparisons (for example `"holm"`, `"bonferroni"`, or `"fdr"`)

### What it does

The function programmatically constructs a model formula from the supplied arguments. If `log_response = TRUE`, the response variable is log-transformed before fitting. For example:

-   one fixed effect: `log(response) ~ factor1 + (1|line)`
-   two fixed effects: `log(response) ~ factor1 * factor2 + (1|line) + (1|line:date)`

The function then:

-   fits the model with `lmer()` from the **lme4** [@lme4] package
-   checks model singularity with `isSingular()`
-   if singularity is detected and `random_effects2` is provided, simplifies the random-effects structure and refits the model using the same fixed-effects formula but with `(1|random_effects2)` as the random effect
-   extracts summaries of the random and fixed effects
-   computes ANOVA statistics using the **lmerTest** [@lmerTest] package, which provides F-tests and p-values using the default Satterthwaite approximation for denominator degrees of freedom
-   computes estimated marginal means and pairwise contrasts for the specified fixed effects and, when applicable, interaction contrasts, using the **emmeans** [@emmeans] package with p-values adjusted by the specified method
-   returns estimated marginal mean predictions and approximate confidence bounds; if the response variable was log-transformed, these predictions are back-transformed to the original response scale for interpretability

### Outputs

The returned object is a list containing:

-   `input_data`: the input data with a combined `condition` column added for downstream matching and plotting
-   `model_formula`: the fitted model formula
-   `anova`: ANOVA table for the fitted model
-   `random_effects`: variance components extracted from the random-effects structure
-   `fixed_effects`: fixed-effect coefficient table from `summary(model)`
-   `emmeans_formula`: formula used to generate estimated marginal means
-   `emmeans_contrasts`: pairwise contrasts for main effects and, when applicable, interactions
-   `emmeans_predictions`: estimated marginal means with back-transformed predictions and confidence bounds

### Function definition

```{r, echo=FALSE, results='asis'}
cat("```r\n")
cat(paste(deparse(mixed_model_with_emmeans), collapse = "\n"))
cat("\n```")
```

## `format_stats_results()`

This function takes the output of `mixed_model_with_emmeans()` and reformats it into tidy tables that are easier to join, filter, annotate, and plot.

### Inputs

-   `stats_results`: the list returned by `mixed_model_with_emmeans()`
-   `grouping_vars`: variables used to summarize the original observed data alongside model predictions. The default is `condition`, a combined variable created in `mixed_model_with_emmeans()`, but additional variables from the original data can be included when needed for downstream plotting or annotation.
-   `p.value_threshold`: threshold used to create a simple significance label column (`sign`); the default is `0.05`

### What it does

This function:

-   summarizes the original observed data by group and calculates the mean and standard error for each group, which can be used for plotting alongside model predictions
-   joins those summaries to the `emmeans` predictions
-   matches each contrast to its corresponding predicted means, allowing calculation of fold changes (FC) and log2 fold changes (log2FC) between contrast groups
-   returns a clean structure for downstream visualization and reporting, including columns named `group1` and `group2` that can be used to annotate significant differences in plots, as well as a simple significance label (`sign`) based on the specified p-value threshold

### Outputs

The returned object is a list containing:

-   `input_data`: the original input data with a combined `condition` column
-   `predictions`: estimated marginal means joined with observed summary statistics
-   `main`: formatted main-effect contrasts with columns for `group1`, `group2`, `FC`, `log2FC`, and `sign`
-   `interaction`: formatted interaction contrasts, if present
-   `anova`: passed through from `mixed_model_with_emmeans()`
-   `random_effects`: passed through from `mixed_model_with_emmeans()`
-   `fixed_effects`: passed through from `mixed_model_with_emmeans()`

### Function definition

```{r, echo=FALSE, results='asis'}
cat("```r\n")
cat(paste(deparse(format_stats_results), collapse = "\n"))
cat("\n```")
```