---
execute:
  echo: true
format:
  html:
    suppress-bibliography: true
---

# R Session Info {.unnumbered}

This document lists all R packages used across analysis notebooks in this repository, and details how to reproduce or create similar analysis environments.

## Package Loading

```{r}
#| warning: FALSE
#| message: FALSE

# Core tidyverse (data handling + plotting)
library(tidyverse)

# Data input/output
library(readxl)      # Read Excel files
library(openxlsx)    # Write Excel files
library(png)         # Read PNG images

# Visualization (ggplot2 ecosystem)
library(ggtext)      # Rich text in ggplot
library(scales)      # Axis scaling
library(ggh4x)       # ggplot extensions
library(ggpubr)      # Publication-ready plots
library(cowplot)     # Plot composition
library(plotly)      # Interactive plots
library(ggVennDiagram) # Venn diagrams
library(grid)        # Low-level graphics

# Statistical modeling
library(lme4)        # Mixed-effects models
library(lmerTest)    # p-values for lme4
library(emmeans)     # Estimated marginal means
library(broom)       # Tidy model outputs
library(mclust)      # Clustering & mixture models

# Tables and reporting
library(kableExtra)  # Styled tables

# Numerical / utility
library(pracma)      # Numerical functions (including AUC)

# Bioinformatics / sequencing
library(Biostrings)  # Sequence analysis
library(DESeq2)      # Differential expression
```

## Session Information

This shows the exact versions of R and all packages used in the analysis.

```{r}
sessionInfo()
```

## Reproducibility and Environment

This repository manages R package versions using [`renv`](https://rstudio.github.io/renv/).

To reproduce the environment for this project:

If using a Windows computer, you must download Rtools45 to "make" packages. https://cran.r-project.org/bin/windows/Rtools/rtools45/rtools.html

```{r}
#renv::init() # Was used to create the initial renv environment and lockfile
#renv::snapshot() # Was used to update the renv.lock file after installing new packages
#renv::restore() # Restores the R environment based on the renv.lock file
#renv::status() #check project status
```

This command uses the `renv.lock` file to install the exact package versions required for the analysis.