---
format:
  html:
    suppress-bibliography: true
---

# CFUs Cocolonization with Cps {.unnumbered}

```{r}
# loading packages and functions
library(tidyverse)
library(kableExtra)
library(ggtext)
library(scales)
library(lme4)
library(lmerTest)
library(emmeans)
library(ggh4x)

source("R/LMM_functions.R")
```

## Data Input 

All data inputs and outputs in this repo are relative to **FOLDER_PATH**. Please, save the provided data frames in RDS format in the `dataframes` folder.

```{r}
# Load the BOX_PATH environment variable
FOLDER_PATH <- Sys.getenv("BOX_PATH") 

# Define subdirectories within the Box folder
dataframes_folder <- file.path(FOLDER_PATH, "dataframes")

outputs_folder <- file.path(FOLDER_PATH, "outputs/CFUs")
if (!dir.exists(outputs_folder)) {
  dir.create(outputs_folder, recursive = TRUE)
}
```

```{r}
# Read the provided RDS file
dataframeID <- "DatCol"
CFUs_HNO_Cps <- readRDS(file.path(dataframes_folder, paste0(dataframeID, "_CFUs_HNO_Cps.rds")))

# Filter the data to include only the 24-hour time point
CFUs_HNO_Cps <- CFUs_HNO_Cps %>% 
  filter(time == "24")
```

Log-transformed CFU/HNO-ALI between groups were compared using a LMM to determine statistical significance with `culture` and `strain` as fixed effects and experimental date as a random effect (line was singular).
All pairwise contrasts were performed and adjusted for multiple comparisons using the Holm method. A `sign` annotation column was created to indicate which comparisons are statistically significant with p<0.05.
Vertical brackets represent model-predicted mean values and confidence intervals (± 2 × SEM), back-transformed to the original response scale. 

```{r}
# Run the mixed model with emmeans contrasts
stats_results_Cps <- mixed_model_with_emmeans(
  input_data = CFUs_HNO_Cps, response = "CFU_HNO", log_response = TRUE, adjust = "holm",
  fixed_effects = c("culture", "strain"), random_effects1 = "date"
)
stats_formated_Cps <- format_stats_results(stats_results_Cps, grouping_vars = c("condition", "culture_pair"))

# Save the stats results
saveRDS(stats_formated_Cps, file.path(outputs_folder, "stats_results_HNO_CFUs_CocultureCps.rds"))
```

```{r}
# Render the results
cat("Model Formula:", format(stats_results_Cps$model_formula))
kable(stats_formated_Cps$random_effects, caption = "Random Effects")
kable(stats_formated_Cps$anova, caption = "ANOVA Table")
kable(stats_formated_Cps$main %>%
    mutate(
      p.value.anova = formatC(p.value.anova, format = "e", digits = 2),
      p.value = formatC(p.value, format = "e", digits = 2)
    ) %>%
    select(effect_name, p.value.anova, contrast, condition1, condition2, FC, log2FC, p.value, sign),
  caption = "Main Contrasts with Predicted Log2FC"
)
```

```{r}
# Function to create a ggplot with jittered points, emmeans predictions with error bars, and faceting. 
plot_Cps_function <- function(formated_stats_results, ymin_exp, ymax_exp, title_x = " ", title_y = " ",
                              x_var, y_var, x_labels, fill_var, shape_var, facet_var, facet_labels = NULL,
                              fill_colors, shape_leves, facet_colors, fill_title, shape_title) {
  plot <- ggplot(data = formated_stats_results$input_data) +
    geom_jitter(
      aes(x = {{ x_var }}, y = {{ y_var }}, fill = {{ fill_var }}, shape = {{ shape_var }}),
      size  = 4, stroke = 0.01, width = 0.2, color = "white") +
    
    geom_point(data = formated_stats_results$predictions, aes(x = {{ x_var }}, y = predicted_mean), shape = 3, size = 1) +
    geom_errorbar(data = formated_stats_results$predictions, aes(x = {{ x_var }},
                                                                 y = predicted_mean,
                                                                 ymin = predicted_lower,
                                                                 ymax = predicted_upper),
                  width = .15) +
    
    facet_grid2(
      cols = vars({{ facet_var }}),
      scales = "free_x", space = "free_x",
      strip = strip_themed(background_x = elem_list_rect(fill = facet_colors)),
      labeller = if (!is.null(facet_labels)) labeller(.cols = facet_labels) else label_value
    ) +
    
    scale_fill_manual(values = fill_colors, name = fill_title) +
    scale_shape_manual(values = shape_leves, name = shape_title) +
    scale_x_discrete(labels = x_labels) +
    
    guides(
      fill  = guide_legend(override.aes = list(shape = 21, color = "white", size = 3, stroke = 2)),
      shape = guide_legend(override.aes = list(fill = "black", size = 3, stroke = 2))) +
    
    scale_y_log10(
      limits = c(10^ymin_exp, 10^ymax_exp),
      breaks = 10^(1:ymax_exp),
      labels = trans_format("log10", math_format(10^.x))) +
    labs(x = title_x, y = title_y) +
    theme_bw() +
    theme(
      panel.grid    = element_blank(),
      panel.spacing = unit(0, "lines"),
      strip.text.x = element_markdown(size = 14, face = "bold", color = "white"),
      text          = element_text(size = 16, color = "black"),
      axis.text.x   = element_markdown(angle = 45, hjust = 1, color = "black"),
      axis.text.y   = element_text(color = "black")
    )
  
  return(plot)
}
```

```{r, fig.width=10, fig.height=6, fig.align="center"}
stats_formated_Cps <- format_stats_results(stats_results_Cps, grouping_vars = c("condition", "culture_pair"))

# Create the plot using the function defined above
plot_CFUs_Cps <- plot_Cps_function(
  formated_stats_results = stats_formated_Cps,
  x_var       = culture_pair, title_x = " ",
  y_var       = CFU_HNO, title_y = "CFUs/HNO-ALI", ymin_exp = 1, ymax_exp = 10,
  fill_var    = strain, fill_colors = c("#5F05AF", "#2FD612"), fill_title = "Condition",
  shape_var   = line, shape_leves = c(21, 22), shape_title = "HNO line",
  facet_var   = strain, facet_colors = c("#5F05AF", "#2FD612"),
  facet_labels = c("JE2" = "parent", "dat" = "<i>dat</i>::Tn"),
  
  x_labels <- c(
    "JE2" = "<b>parent</b>",
    "dat" = "<b><i>dat</i>::Tn</b>",
    "JE2_Cps" = "<b>parent + <span style='color:#CC7722'><i>Cps</i><span></b>",
    "dat_Cps" = "<b><i>dat</i>::Tn + <span style='color:#CC7722'><i>Cps</i><span></b>")
)

plot_CFUs_Cps <- plot_CFUs_Cps +
  geom_vline(xintercept = c(1.5), color = "grey70", linewidth = 0.5, linetype = "dashed") +
  annotate("text", x = 0.95, y = Inf, label = "monocolonization", vjust = 2, fontface = "bold") +
  annotate("text", x = 2, y = Inf, label = "cocolonization", vjust = 2, fontface = "bold", color = "#CC7722")

plot_CFUs_Cps

# Save the plot
ggsave(filename = file.path(outputs_folder, "HNO_CFUs_CocultureCps.png"), plot = plot_CFUs_Cps, width = 10, height = 6, dpi = 300)
saveRDS(plot_CFUs_Cps, file.path(outputs_folder, "HNO_CFUs_CocultureCps.rds"))
```

