---
format:
  html:
    suppress-bibliography: true
---

# Growth Curves {.unnumbered}

```{r}
# loading packages
library(tidyverse)
library(ggtext)
library(kableExtra)
library(pracma)
library(lme4)
library(lmerTest)
library(emmeans)

source("R/LMM_functions.R")
```

## Data Input 

All data inputs and outputs in this repo are relative to **FOLDER_PATH**. Please, save the provided data frames in RDS format in the `dataframes` folder.

```{r}
# Load the BOX_PATH environment variable
FOLDER_PATH <- Sys.getenv("BOX_PATH") 

# Define subdirectories within the Box folder
dataframes_folder <- file.path(FOLDER_PATH, "dataframes")

outputs_folder <- file.path(FOLDER_PATH, "outputs/GC_OD600")
if (!dir.exists(outputs_folder)) {
  dir.create(outputs_folder, recursive = TRUE)
}
```

```{r}
# Read the provided RDS file
dataframeID <- "DatCol"
GC_OD600_data <- readRDS(file.path(dataframes_folder, paste0(dataframeID, "_GC_OD600.rds")))
```

```{r}
# Filter the data to include only time points up to 20 hours and exclude the Uncolonized control
GC_OD600_20h <- GC_OD600_data %>%
  filter(time <= 20) %>%
  filter(label != "NB")

# Summarize data by calculating mean and SEM for each time point and condition
GC_OD600_20h_summarized <- GC_OD600_20h %>%
  group_by(label, media, time, strain, vector) %>%
  summarise(
    mean_OD600 = mean(OD600),
    sem_OD600 = sd(OD600) / sqrt(n()),  
    .groups = "drop"
  ) 
```

```{r}
GC_plot_function <- function(data, media_filter, ymin = 0, ymax = 1.8) {
  # Filter data based on date and media type
  filtered_data <- data %>%
    filter(media %in% media_filter) %>%
    droplevels()
  
  # Create the growth curve plot
  growth_curve_plot <- ggplot(filtered_data, aes(x = time, y = mean_OD600, color = strain, shape = vector)) +
    geom_line() +
    geom_point(size = 3) +
    
    scale_color_manual(
      values = c(
        Parent  = "#5F05AF",
        "dat::Tn"  = "#2FD612",
        Δalr1 = "#03C0FF"
      ),
      labels = c(
        Parent = "<b><span style='color:#5F05AF'>parent<span></b>",
        "dat::Tn" = "<b><span style='color:#2FD612'><i>dat</i>::Tn<span></b>",
        "Δalr1" = "<b><span style='color:#03C0FF'>Δ  <i>alr1</i><span></b>"
      )
    ) +
    scale_shape_manual(
      values = c(
        None     = 16,  # filled circle
        EV     = 1,   # empty circle
        "PpepV dat"  = 17,  # filled triangle
        "Pdat dat"   = 2,   # empty triangle
        alr1  = 15   # filled square
      ),
      labels = c(
        None = "None",
        EV = "EV",
        "PpepV dat" = "P<i><sub>pepV</sub>dat</i>",
        "Pdat dat" = "P<i><sub>dat</sub>dat</i>",
        alr1 = "P<i><sub>alr1</sub>alr1</i>"
      )
    ) +
    
    geom_errorbar(aes(ymin = mean_OD600 - sem_OD600, ymax = mean_OD600 + sem_OD600), width = 0.2) +
    
    labs(title = paste(paste(media_filter)),
         x = "Time (hours)",
         y = "Mean OD<sub>600</sub> ± SEM") +
    
    scale_y_continuous(limits = c(ymin, ymax), expand = c(0, 0)) +
    theme_bw() +
    theme(
      plot.title = element_text(vjust = -10, hjust = 0.05, size = 14, face = "bold"),
      legend.text = element_markdown(size = 12),
      legend.title = element_markdown(size = 12),
      panel.grid = element_blank(),
      text = element_text(size = 16, color = "black"),
      axis.text = element_markdown(color = "black"),
      axis.title.y = element_markdown()
    ) +
    
    guides(color = guide_legend(override.aes = list(linetype = 0, shape = NA), keywidth = 0))                               
  
  return(growth_curve_plot)
}
```

## Main figure 

```{r, fig.width=10, fig.height=6, fig.align="center"}
figMain <- GC_plot_function(GC_OD600_20h_summarized %>% filter (vector != "None"), media_filter = c("CDM19 - ala"), ymin = 0, ymax = 1.5)
figMain

# Save the plot
ggsave(filename = file.path(outputs_folder, "figMain.png"), plot = figMain, width = 10, height = 6, dpi = 300)
saveRDS(figMain, file.path(outputs_folder, "figMain.rds"))
```

## Stats and Supplemental Figures

The Area Under the Curve (AUC) was calculated for each growth curve to quantitatively compare the growth across different conditions. The AUC was computed using the trapezoidal rule from the `pracma` package.

```{r}
# Calculate AUC for each growth curve using package pracma's trapz function, which performs numerical integration using the trapezoidal rule.
GC_AUC <- GC_OD600_20h %>%
  mutate(
    media = recode(media, "CDM19 - ala" = "CDM19Ala"),
    across(where(is.character), ~ str_replace_all(.x, " ", "")),
    across(where(is.factor), ~ factor(str_replace_all(as.character(.x), " ", "")))
  ) %>%
  group_by(label, media, strain, vector, date) %>%  
  summarise(AUC = trapz(time, OD600), .groups = "drop")

# Make a dataframe for each media condion for stats 
GC_AUC_CDM19Ala <- GC_AUC %>%
  filter(media == "CDM19Ala") %>%
  droplevels()

GC_AUC_TSB <- GC_AUC %>%
  filter(media == "TSB") %>%
  droplevels()

GC_AUC_CDM20 <- GC_AUC %>%
  filter(media == "CDM20") %>%
  droplevels()
```

Data for each media conditions were analyzed separately using a LMM to determine statistical significance with `label` as fixed effects and experimental date as a random effect.
We analyzed pairwise contrasts between Parent with the EV or no vector compared to all other strains. Also we look at all contrast within the strains dat::Tn and Δalr1.
We then adjusted the p-values for multiple comparisons using the Holm method and created a `sign` annotation column to indicate which comparisons are statistically significant with p<0.05. 

### TSB

```{r}
# Run the mixed model with emmeans contrasts for the JE2 dataset.
stats_results_AUC_TSB <- mixed_model_with_emmeans(
  input_data = GC_AUC_TSB, response = "AUC", log_response = FALSE, adjust = "none",
  fixed_effects = c("label"), random_effects1 = "date"
)
stats_formated_AUC_TSB <- format_stats_results(stats_results_AUC_TSB)
stats_formated_AUC_TSB$main = stats_formated_AUC_TSB$main %>%
  filter(condition1 == "Parent" | condition2 == "Parent" | condition1 == "Parent_EV" | condition2 == "Parent_EV" 
         | (grepl("dat::Tn", condition1) & grepl("dat::Tn", condition2))
         | (grepl("Δalr1", condition1) & grepl("Δalr1", condition2))
  ) %>%
  mutate(
    p.value.adj = p.adjust(p.value, method = "holm"),
    sign = case_when(p.value.adj < 0.05 ~ "*", TRUE ~ "")
  )

# Save the stats results
saveRDS(stats_formated_AUC_TSB, file.path(outputs_folder, "stats_results_AUC_TSB.rds"))
```

```{r}
# Render the results
cat("Model Formula:", format(stats_results_AUC_TSB$model_formula))
kable(stats_formated_AUC_TSB$random_effects, caption = "Random Effects")
kable(stats_formated_AUC_TSB$anova, caption = "ANOVA Table")
kable(stats_formated_AUC_TSB$main %>%
    mutate(
      p.value.anova = formatC(p.value.anova, format = "e", digits = 2),
      p.value = formatC(p.value, format = "e", digits = 2),
      p.value.adj = formatC(p.value.adj, format = "e", digits = 2)
    )  %>%
    select(effect_name, p.value.anova,  condition1, condition2, predicted_mean.1, predicted_mean.2, FC, log2FC, p.value, p.value.adj, sign),
  caption = "Main Contrasts with Predicted Log2FC"
)
```

```{r, fig.width=10, fig.height=6, fig.align="center"}
figSuppA <- GC_plot_function(GC_OD600_20h_summarized, media_filter = c("TSB"), ymin = 0, ymax = 1.5)
figSuppA

# Save the plot
ggsave(filename = file.path(outputs_folder, "figSuppA.png"), plot = figSuppA, width = 10, height = 6, dpi = 300)
saveRDS(figSuppA, file.path(outputs_folder, "figSuppA.rds"))
```

### CDM20

```{r}
# Run the mixed model with emmeans contrasts for the JE2 dataset.
stats_results_AUC_CDM20 <- mixed_model_with_emmeans(
  input_data = GC_AUC_CDM20, response = "AUC", log_response = FALSE, adjust = "none",
  fixed_effects = c("label"), random_effects1 = "date"
)
stats_formated_AUC_CDM20 <- format_stats_results(stats_results_AUC_CDM20)
stats_formated_AUC_CDM20$main = stats_formated_AUC_CDM20$main %>%
  filter(condition1 == "Parent" | condition2 == "Parent" | condition1 == "Parent_EV" | condition2 == "Parent_EV" 
         | (grepl("dat::Tn", condition1) & grepl("dat::Tn", condition2))
         | (grepl("Δalr1", condition1) & grepl("Δalr1", condition2))
  ) %>%
  mutate(
    p.value.adj = p.adjust(p.value, method = "holm"),
    sign = case_when(p.value.adj < 0.05 ~ "*", TRUE ~ "")
  )

# Save the stats results
saveRDS(stats_formated_AUC_CDM20, file.path(outputs_folder, "stats_results_AUC_CDM20.rds"))
```

```{r}
# Render the results
cat("Model Formula:", format(stats_results_AUC_CDM20$model_formula))
kable(stats_formated_AUC_CDM20$random_effects, caption = "Random Effects")
kable(stats_formated_AUC_CDM20$anova, caption = "ANOVA Table")
kable(stats_formated_AUC_CDM20$main %>%
    mutate(
      p.value.anova = formatC(p.value.anova, format = "e", digits = 2),
      p.value = formatC(p.value, format = "e", digits = 2),
      p.value.adj = formatC(p.value.adj, format = "e", digits = 2)
    )  %>%
    select(effect_name, p.value.anova,  condition1, condition2, predicted_mean.1, predicted_mean.2, FC, log2FC, p.value, p.value.adj, sign),
  caption = "Main Contrasts with Predicted Log2FC"
)
```

```{r, fig.width=10, fig.height=6, fig.align="center"}
figSuppB <- GC_plot_function(GC_OD600_20h_summarized, media_filter = c("CDM20"), ymin = 0, ymax = 1.5)
figSuppB

# Save the plot
ggsave(filename = file.path(outputs_folder, "figSuppB.png"), plot = figSuppB, width = 10, height = 6, dpi = 300)
saveRDS(figSuppB, file.path(outputs_folder, "figSuppB.rds"))
```

### CDM19 - ala

```{r}
# Run the mixed model with emmeans contrasts for the JE2 dataset.
stats_results_AUC_CDM19Ala <- mixed_model_with_emmeans(
  input_data = GC_AUC_CDM19Ala, response = "AUC", log_response = FALSE, adjust = "none",
  fixed_effects = c("label"), random_effects1 = "date"
)
stats_formated_AUC_CDM19Ala <- format_stats_results(stats_results_AUC_CDM19Ala)
stats_formated_AUC_CDM19Ala$main = stats_formated_AUC_CDM19Ala$main %>%
  filter(condition1 == "Parent" | condition2 == "Parent" | condition1 == "Parent_EV" | condition2 == "Parent_EV" 
         | (grepl("dat::Tn", condition1) & grepl("dat::Tn", condition2))
         | (grepl("Δalr1", condition1) & grepl("Δalr1", condition2))
  ) %>%
  mutate(
    p.value.adj = p.adjust(p.value, method = "holm"),
    sign = case_when(p.value.adj < 0.05 ~ "*", TRUE ~ "")
  )

# Save the stats results
saveRDS(stats_formated_AUC_CDM19Ala, file.path(outputs_folder, "stats_results_AUC_CDM19Ala.rds"))
```

```{r}
# Render the results
cat("Model Formula:", format(stats_results_AUC_CDM19Ala$model_formula))
kable(stats_formated_AUC_CDM19Ala$random_effects, caption = "Random Effects")
kable(stats_formated_AUC_CDM19Ala$anova, caption = "ANOVA Table")
kable(stats_formated_AUC_CDM19Ala$main %>%
    mutate(
      p.value.anova = formatC(p.value.anova, format = "e", digits = 2),
      p.value = formatC(p.value, format = "e", digits = 2),
      p.value.adj = formatC(p.value.adj, format = "e", digits = 2)
    )  %>%
    select(effect_name, p.value.anova,  condition1, condition2, predicted_mean.1, predicted_mean.2, FC, log2FC, p.value, p.value.adj, sign),
  caption = "Main Contrasts with Predicted Log2FC"
)
```

```{r, fig.width=10, fig.height=6, fig.align="center"}
figSuppC <- GC_plot_function(GC_OD600_20h_summarized, media_filter = c("CDM19 - ala"), ymin = 0, ymax = 1.5)
figSuppC

# Save the plot
ggsave(filename = file.path(outputs_folder, "figSuppC.png"), plot = figSuppC, width = 10, height = 6, dpi = 300)
saveRDS(figSuppC, file.path(outputs_folder, "figSuppC.rds"))
```

